Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms

### 2023-06-15 output/latex/manuscript.tex
update latex

\begin{figure}
\hypertarget{fig:entire_process}{%
\centering
\includesvg[width=1\textwidth,height=\textheight]{images/entire_process/entire_process.svg}
\caption{\textbf{Schematic of the PhenoPLIER framework.}
\textbf{a)} High-level schematic of PhenoPLIER (a gene module-based method) in the context of TWAS (single-gene) and GWAS (single-variant).
In GWAS, we identify variants associated with traits.
In TWAS, first, we identify variants that are associated with gene expression levels (eQTLs); then, prediction models based on eQTLs are used to impute gene expression, which is used to compute gene-trait associations.
Resources such as LINCS L1000 provide information about how a drug perturbs gene expression; at the bottom-right corner, we show how a drug downregulates two genes (A and C).
In PhenoPLIER, these data types are integrated using groups of genes co-expressed across one or more conditions (such as cell types) that we call gene modules or latent variables/LVs.
\textbf{b)} The integration process in PhenoPLIER uses low-dimensional representations (matrices \(\mathbf{Z}\) and \(\mathbf{B}\)) learned from large gene expression datasets (top).
We used gene-drug information \(\mathbf{L}\) from LINCS L1000 and gene-trait associations \(\mathbf{M}\) from TWAS: PhenomeXcan was used as the discovery cohort, and eMERGE as replication (middle).
PhenoPLIER provides three computational components (bottom):
1) an LV-based regression model that associates an LV \(j\) (\(\mathbf{Z}_j\)) with a trait \(i\) (\(\mathbf{M}_i\)),
2) a clustering framework that learns groups of traits from TWAS associations projected into the LV space (\(\hat{\mathbf{M}}\)),
and 3) an LV-based drug repurposing approach that uses the projection of TWAS (\(\hat{\mathbf{M}}\)) and LINCS L1000 (\(\hat{\mathbf{L}}\)) into the LV space.
\textbf{c)} Genes that are part of LV603, termed as a neutrophil signature\textsuperscript{\protect\hyperlink{ref-14rnBunuZ}{44}}, were expressed in relevant cell types (top).
LV603 was associated in PhenoPLIER with neutrophil counts and other white blood cells (bottom, showing the top 10 traits for LV603 after projecting gene-trait associations in PhenomeXcan).
eQTLs: expression quantitative trait loci;
MVN: multivariate normal distribution;
PBMC: peripheral blood mononuclear cells;
mDCs: myeloid dendritic cells.}\label{fig:entire_process}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:drug_disease:roc_pr}{%
\centering
\includesvg[width=0.8\textwidth,height=\textheight]{images/drug_disease_prediction/roc_pr_curves.svg}
\caption{\textbf{Drug-disease prediction performance for gene-based and LV-based approaches.}
The receiver operating characteristic (ROC) (left) and the precision-recall curves (right) for a gene-based and LV-based approach.
AUC: area under the curve; AP: average precision.}\label{fig:drug_disease:roc_pr}
}
\end{figure}

>>>>

\begin{figure}
\hypertarget{fig:entire_process}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/entire_process/entire_process.pdf}
\caption{\textbf{Schematic of the PhenoPLIER framework.}
\textbf{a)} High-level schematic of PhenoPLIER (a gene module-based method) in the context of TWAS (single-gene) and GWAS (single-variant).
In GWAS, we identify variants associated with traits.
In TWAS, first, we identify variants that are associated with gene expression levels (eQTLs); then, prediction models based on eQTLs are used to impute gene expression, which is used to compute gene-trait associations.
Resources such as LINCS L1000 provide information about how a drug perturbs gene expression; at the bottom-right corner, we show how a drug downregulates two genes (A and C).
In PhenoPLIER, these data types are integrated using groups of genes co-expressed across one or more conditions (such as cell types) that we call gene modules or latent variables/LVs.
\textbf{b)} The integration process in PhenoPLIER uses low-dimensional representations (matrices \(\mathbf{Z}\) and \(\mathbf{B}\)) learned from large gene expression datasets (top).
We used gene-drug information \(\mathbf{L}\) from LINCS L1000 and gene-trait associations \(\mathbf{M}\) from TWAS: PhenomeXcan was used as the discovery cohort, and eMERGE as replication (middle).
PhenoPLIER provides three computational components (bottom):
1) an LV-based regression model that associates an LV \(j\) (\(\mathbf{Z}_j\)) with a trait \(i\) (\(\mathbf{M}_i\)),
2) a clustering framework that learns groups of traits from TWAS associations projected into the LV space (\(\hat{\mathbf{M}}\)),
and 3) an LV-based drug repurposing approach that uses the projection of TWAS (\(\hat{\mathbf{M}}\)) and LINCS L1000 (\(\hat{\mathbf{L}}\)) into the LV space.
\textbf{c)} Genes that are part of LV603, termed as a neutrophil signature\textsuperscript{\protect\hyperlink{ref-14rnBunuZ}{44}}, were expressed in relevant cell types (top), with 53 independent samples expressed in Neutrophils, 59 in Granulocytes, and 20 in Whole blood, 56 in PBMC, 8 in mDCs, 29 in Monocytes, and 5 in Epithelial cells (the boxplot shows the 25th, 50th and 75th percentiles while the whiskers extend to the minimum/maximum values).
LV603 was associated in PhenoPLIER with neutrophil counts and other white blood cells (bottom, showing the top 10 traits for LV603 after projecting gene-trait associations in PhenomeXcan).
eQTLs: expression quantitative trait loci;
MVN: multivariate normal distribution;
PBMC: peripheral blood mononuclear cells;
mDCs: myeloid dendritic cells.}\label{fig:entire_process}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:drug_disease:roc_pr}{%
\centering
\includesvg[width=0.8\textwidth,height=\textheight]{images/drug_disease_prediction/roc_pr_curves.svg}
\caption{\textbf{Drug-disease prediction performance for gene-based and LV-based approaches.}
The receiver operating characteristic (ROC) (left) and the precision-recall curves (right) for a gene-based and LV-based approach.
``Random'' refers to the average precision of a hundred classifiers with randomly permuted scores, where the error band represents the 95\% confidence interval.
AUC: area under the curve; AP: average precision.}\label{fig:drug_disease:roc_pr}
}
\end{figure}
